---
title: "Class 7 - R Notebooks & Statistics in R"
output:
  html_notebook: default
---

```{r}
rm(list=ls())

library(tidyverse)

## import data

# maryland report cards
msde_report_cards <- read_csv("data/2019_Accountability_Schools.csv")

# ccd direct certification
ccd_direct_cert <- read_csv("data/ccd_sch_033_1819_l_1a_091019.csv", 
                            col_types = cols(
                              SCHOOL_YEAR = col_character(),
                              FIPST = col_character(),
                              STATENAME = col_character(),
                              ST = col_character(),
                              SCH_NAME = col_character(),
                              STATE_AGENCY_NO = col_character(),
                              UNION = col_character(),
                              ST_LEAID = col_character(),
                              LEAID = col_character(),
                              ST_SCHID = col_character(),
                              NCESSCH = col_character(),
                              SCHID = col_character(),
                              DATA_GROUP = col_character(),
                              LUNCH_PROGRAM = col_character(),
                              STUDENT_COUNT = col_double(),
                              TOTAL_INDICATOR = col_character(),
                              DMS_FLAG = col_character()
                              )
                            )
```

### MSDE Report Card

- state_schid : "MD-" + LSS Number {2 digits} + "-" + LSS Number {2 digits} + School Number {4 digits}
- school_name : School Name
- star_rating : Star Rating

Observations - All Baltimore City Schools : LSS Number == 30

```{r}
msde_report_cards <- msde_report_cards %>%
  filter(`LSS Number` == 30) %>%
  mutate(
    lss_number_f <- sprintf("%02i", `LSS Number`),
    school_number_f <- sprintf("%04i", `School Number`),
    state_schid = paste0("MD-", lss_number_f, "-", lss_number_f, school_number_f),
  ) %>%
  rename(school_name = `School Name`, star_rating = `Star Rating`) %>%
  select(state_schid, school_name, star_rating)
```

### CCD Direct Certification

- state_schid : ST_SCHID
- school_name : SCH_NAME
- direct_cert_per : Calculated after reshape to wide { direct_cert_num / total_enrollment * 100 }

Observations - All Baltimore City Schools : ST_LEAID == "MD-30"

```{r}
ccd_direct_cert <- ccd_direct_cert %>%
  filter(
    LUNCH_PROGRAM == "Free lunch qualified" | LUNCH_PROGRAM == "Not Applicable",
    ST_LEAID == "MD-30"
  ) %>%
  select(ST_SCHID, SCH_NAME, DATA_GROUP, STUDENT_COUNT) %>%
  pivot_wider(
    c(ST_SCHID, SCH_NAME),
    names_from = DATA_GROUP,
    values_from = STUDENT_COUNT
  ) %>%
  magrittr::set_colnames(c("state_schid", "school_name", "total_enrollment", "direct_cert_num")) %>%
  mutate(direct_cert_per = direct_cert_num / total_enrollment * 100) %>%
  select(-total_enrollment, -direct_cert_num)
```

### Create Analytic

```{r}
analytic <- left_join(
  ccd_direct_cert,
  msde_report_cards,
  by = "state_schid"
)
```

## Analyses

Correlation

```{r}
cor(analytic$star_rating, analytic$direct_cert_per, use = "complete.obs")
```

Linear Model

```{r}
fit <- lm(star_rating ~ direct_cert_per, data=analytic)

print(fit)
```

```{r}
print(summary(fit))
```

```{r}
formula <- as.formula("star_rating ~ direct_cert_per")

fit <- lm(formula, data=analytic)

print(fit)
```

```{r}
library(broom)

tidy(fit)
```

```{r}
augment(fit)
```

```{r}
glance(fit)
```

```{r}
library(knitr)
library(kableExtra)

tidy(fit) %>% kable %>% kable_styling
```

```{r, results='asis'}
library(stargazer)

r <- stargazer(fit, type="html")

suppressWarnings(print(r))
```

