---
title: "Potential Outcomes in R"
output: html_notebook
---

```{r}
rm(list=ls())

library(tidyverse)
library(latex2exp)

# setup libraries
theme_set(theme_minimal())

# functions
constrain <- function(v, min, max) {
  case_when(
    v < min ~ min,
    v > max ~ max,
    TRUE ~ v
  )
}
```

## Potential Outcomes

$$
ATE = E[Y_i^1 - Y_i^0]
$$

#### Task 1: Simulate 100,000 student test scores with a mean of 100 and standard deviation of 25. Test scores should not be lower than 0 or greater than 200. Then, plot the distribution. (Hint: Look up the R stats function rnorm.)


```{r}
N <- 100000
mean <- 100
sd <- 25

Y0 <- rnorm(N, mean, sd) %>% constrain(0, 200)

ggplot() +
  geom_density(aes(x=Y0)) +
  labs(x=TeX("$Y^0$"))
```
#### Task 2: Imagine another universe where the only difference is the students from Task 1 had a treatment which boosted their performance on the test. Add a mean shift of 5 with a standard deviation of 2 to all the test scores from Task 1. Test scores should still not be lower than 0 or greater than 200. Then, plot both distributions together.

```{r}
Y1 <- (Y0 + rnorm(N, 5, 2)) %>% constrain(0, 200)

ggplot() +
  geom_density(aes(x=Y0), color="lightgrey") +
  geom_density(aes(x=Y1)) +
  labs(x=TeX("$Y^1$"), caption=TeX("The grey line is $Y^0$."))
```
#### Task 3: Calculate the true treatment effect for each student. Then, calculate the true average treatment effect for the population. Plot the treatment effect distribution and the average treatment effect.

```{r}
TE <- Y1 - Y0
ATE <- mean(TE)

ggplot() +
  geom_density(aes(x = TE)) +
  geom_vline(aes(xintercept = ATE), color="red", linetype="dashed") +
  geom_text(aes(x=6.6, y=0.04, label=paste0("ATE = ", round(ATE, 4))), color="red") +
  labs(x = TeX("$Y^1 - Y^0$"))
```
Unfortunately, $Y^0$ and $Y^1$ can never be observed for the same individual. So, we have to estimate. Since,

$$
\begin{align}
ATE &= E[Y_i^1 - Y_i^0] \\ 
&= E[Y^1] - E[Y^0]
\end{align}
$$
the outcome under treatment and control can be estimated separately from samples of the population.

$$
E[Y^1|S=1] = E[Y^1] \\
E[Y^0|S=2] = E[Y^0]
$$
The estimated ATE then is given by:

$$
\begin{align}
ATE &= E[Y_i^1 - Y_i^0] \\
&= E[Y^1] - E[Y^0] \\
&= E[Y^1|S=1] - E[Y^0|S=2]
\end{align}
$$

```{r}
n <- 200

population <- data.frame(Y0, Y1, TE)
sample <- slice_sample(population, n=n)

data.frame(
  variable = c("Y0", "Y1"),
  pop_mean = c(mean(population$Y0), mean(population$Y1)),
  sample_mean = c(mean(sample$Y0), mean(sample$Y1)),
  est_se = c(sd(sample$Y0)/sqrt(n), sd(sample$Y1)/sqrt(n)),
  actual_se = c(sd(population$Y0)/sqrt(n), sd(population$Y1)/sqrt(n))
)
```

$$
se = \frac{sd}{\sqrt{n}}
$$
What is the standard error?

#### Task 4: Demonstrate that the standard error is the standard deviation of the mean upon repeated sampling.

```{r}
n <- 200

sample_means <- sapply(1:10000, function(i) {
  sample <- slice_sample(population, n=n)
  sample_mean <- mean(sample$Y1)
  
  return(sample_mean)
})

ggplot() +
  geom_density(aes(x = sample_means)) +
  annotate("text", 109, 0.125, label=paste0("SD = ", round(sd(sample_means), 4)))
```
$$
ATE = E[Y_i^1 - Y_i^0] = E[Y^1] - E[Y^0] = E[Y^1|S=1] - E[Y^0|S=2]
$$
```{r}
sample_1 <- slice_sample(population, n=n)
sample_2 <- slice_sample(population, n=n)

t.test(sample_1$Y1, sample_2$Y0, paired=T, var.equal=T)

df <- data.frame(
  Y = c(sample_1$Y1, sample_2$Y0),
  D = rep(c(1, 0), each=n)
)

lm(Y ~ D, data=df) %>% summary
```
That's not actually how an RCT is done.

$$
ATE = E[Y_i^1 - Y_i^0] = E[Y^1] - E[Y^0] = E[Y^1|S=1, D=1] - E[Y^0|S=1, D=0]
$$
```{r}
sample_RCT <- slice_sample(population, n=n*2)

sample_RCT <- sample_RCT %>%
  mutate(
    D = ifelse(row_number() <= n, 1, 0),
    Y = ifelse(D == 1, Y1, Y0)
    )

t.test(sample_RCT[which(sample_RCT$D == 1), "Y"], sample_RCT[which(sample_RCT$D == 0), "Y"], paired=T, var.equal=T)

sample_RCT %>%
  lm(Y ~ D, data=.) %>% summary
```
These aren't actually good estimators for an RCT because they ignore the design. Randomization in an RCT is simple random sampling from a finite sample not an infinite population.

### Randomization test

```{r}
treatment_effect_rct <- function(df) {
  Y1 <- filter(df, D==1) %>% pull(Y)
  Y0 <- filter(df, D==0) %>% pull(Y)
  
  mean(Y1) - mean(Y0)
}

shuffle_treatment_rct <- function(df) {
  reshuffled_df <- df
  reshuffled_df[,"D"] <- sample(reshuffled_df[,"D"], size = nrow(reshuffled_df), replace = FALSE)
  return(reshuffled_df)
}

actual_treatment_effect <- treatment_effect_rct(sample_RCT)

random_treatment_effects <- sapply(1:10000,
  function(x) {
    df <- shuffle_treatment_rct(sample_RCT)
    return(treatment_effect_rct(df))
  }
)

p <- sum(random_treatment_effects > actual_treatment_effect) / length(random_treatment_effects)

ggplot() +
  geom_histogram(aes(x = random_treatment_effects), bins=40) +
  geom_vline(aes(xintercept = actual_treatment_effect), color="red", linetype="dashed") +
  annotate("text", 8, 600, label=paste0("p = ", p))
```






